---
title     : A Machine Learning Model to Predict Self Movement 
subtitle  : part of Practical Machine Learning in Coursera
author    : Juan Carlos Garcia
date      : "Friday, August 22, 2014"
output    : html_document
---

```{r lib,  echo=FALSE, results='hide'}
library(caret)
library(lattice)
library(ggplot2)
library(randomForest)
```

A Machine Learning Model to Predict Self Movement 
=================================================

## Executive Summary
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible quantify self movement to improve health by finding patterns to predict successfully whether barbell lifts are done correctly.

## Data Preparation
```{r data,  echo=FALSE, results='hide'}

rawTraining <- read.csv('pml-training.csv') 
rawValidation <- read.csv('pml-testing.csv')

nzvData <- nearZeroVar(rawValidation)
rawTraining <- rawTraining[,-nzvData]
rawTraining <- rawTraining[,-(1:6)]
dim(rawTraining)

```
The process to prepare the data is:
* Load data into two raw datasets
* Remove missing data using caret nearZeroVar()
* Remove 'X' that is a sequence for each data row that might cause bias.
* At this point we have 52 predictors plus the outcome variable.
Now we are ready to partition the data by 60/40 ratio to build our model.
```{r partition, echo=FALSE}
partData <- createDataPartition(y=rawTraining$classe,p=0.6,list=F)
training <- rawTraining[partData,]
testing <- rawTraining[-partData,]
print('training')
dim(training)
print('testing')
dim(testing)
colnames(rawTraining)
```
---
## Feature and Model Selection
After evaluate several models using caret package in R like and evaluate the accuracy the results determine the random forest got the best accuracy (rf, 99%) compared with other models like k-nearest neighborhood (knn, 73%), bootstrapped random forest (rpart, 87%).

The Random Forest method takes longer but 

```{r model, echo=FALSE, cache=TRUE}
tc = trainControl(method="cv",number=4)
fit <- train(classe ~ .,data=training, method="rf",trControl=tc)
fit$finalModel
```
## Cross Validation
```{r predict, echo=FALSE}
predictResult <- predict(fit,testing)
qplot(predictResult,classe,data=testing,main="Cross Validation",xlab="Predict",ylab="Actual")
confusionMatrix(predictResult, testing$classe)
```
## Variable Importance
This section use varImp() to determine what variables are important for the model.
```{r varImp, echo=FALSE}
varImp(fit)
```
## Prediction Assigment
This is the prediction and export of the Prediction Assignment that is submitted in the Coursera.
```{r assigment}
predictAssigment <- predict(fit, newdata=rawValidation)
qplot(rawValidation$problem_id,predictAssigment, colour=predictAssigment,xlab="Problem id",ylab="Predict classe", height="100")

#Convert data frame to character vector
answers <- data.frame(lapply(predictAssigment, as.character), stringsAsFactors=FALSE)

#Function to output validation set answers in submittable format
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answers)
```

